Sketch Recogniton using Particle Swarm Algorithm

### 2009-04-09 draft/ChapterIntroduction.tex
now it 9 - appril and after club day

%\usepackage{graphicx}
\chapter {Introduction}
\pagenumbering{arabic}

Computer research nowadays aimed at facilitating computer human interaction in every way. Pen-based interfaces give the user a pencil-paper like feeling that enhances interaction more than the current used keyboard-mouse computer interfaces. However, until now there are hardly any complete pen based computer systems. Generally, Current Pen-Based systems make use of the Pen or stylus to perform the same role as the mouse / keyboard. A minority of such systems are equipped with handwriting or digit recognition that is typically restricted for specific language.

@@

Sketchs here stand for any hand drawn drawing the user draws using digital stylus or similar device. Normally, sketchs are composed of a set of strokes. The time between pen-up and pen-down events identify each stroke. The strokes are mainly the path of points, which the user path through between the pen-down and pen-up events while using the stylus or similar device.

%Sketch recognition is procedure that will input a sequence of points from user then segment them into clusters of strokes and recognize those clusters to primitive geometric objects and shapes. However, Sketch understanding is transforming the sketch from a group of primitive shapes into comprehendible identified symbols consistent with each other.
% the entire sketched symbols.
%Sketch recognition is defined as the process of identifying symbols that users draw using single or multiple strokes. Users draw strokes using a pen and the system should immediately interprets their strokes into objects that can be easily manipulated. This paper uses Particle Swarm Algorithm (PSO) to divide the strokes the user draws into meaningful geometric primitives. These geometric primitives are grouped to formulate symbols which are further identified. The results show that using PSO improves segmentation results which guide the symbol recognition phase. This paper uses Support Vector Machine(SVM) classifier which further improves the final recognition accuracy.  %u Final recognition has improved accuracy after the PSO algorithm using a SVM classifier.  
 %Hand drawn sketches are the natural method of exchanging thoughts between engineers and humans in general. Nevertheless, researchers barely give attention to improve current gesture and sketch recognition systems. In a sketch understanding system, users need to feel the freedom of paper and pencil along with the flexibility of computer manipulation.

 %A picture speak a thousand words is also true for sketches. 
%Scientists generally and engineers specifically express thoughts and designs using sketches. Engineers use sketches to exchange designs as a natural method of communication rather than writing or speaking. Engineers use paper and pencil design in early stages of design. Design engineers need a powerful computer based system with symbol recognition to help them design, manipulate and store sketches more effectively than using only papers. Increasing the interaction between computers and users in sketch and CAD systems has been the reason for the emerging of few advanced sketch recognition systems \cite{CAD}. Sketch recognition system converts the drawings and informal graphs user draws into objects and symbols that can be used in further interaction with the system.

%Sketch recognition is divided into four main steps preprocessing, segmentation, symbol recognition and finally sketch understanding. The preprocessing process saves user input stroke points from the hardware and collects basic information about the stroke then proceeds to remove noise and primary data calculation. In segmentation phase, the system divides the stroke into a set of simple geometrical primitives or segments. The third phase, sketch recognition, is clustering strokes and segments to formulate symbols that can be recognized by a classifier system. Finally, the sketch understanding phase uses priori information about the sketch to support identification of symbols by excluding any possible invalid symbols. 
%the basic symbols are identified in consistent with the whole sketch. \\
%from the user information from the users converts the strokes the user draw into segments which is recognized into basic symbols that can be easily manipulated. Finally, the basic symbols are interpreted in consistent with the whole sketch. \\
%This paper introduces a sketch recognition system. The presented system uses' particle swarm algorithm (PSO) to optimally segment users strokes. Users can draw symbols using any number of strokes in any order they like. The system segments stroke user draws then cluster symbols together then passes them to the SVM symbol recognizer. The experiments in section \ref{sec:Experiments} shows that the proposed system improves the final symbol recognition accuracy.

%The remaining of this papers is organized as follows; section \ref{Sec:RelatedWork} provides a brief literature review.  Section \ref{PSO} explains the general particle swarm algorithm. The proposed system block diagram along with a detailed system description is presented in Section \ref{Sysdisc}. The experiments and results achieved are explained in Section \ref{sec:Experiments}. Finally, section \ref{ConclusionandFutureWork} states the conclusion and future work. 
%The next section will relate previous work in the field. The remaining of the paper will be as following; Section \ref{PSO} will explain the general particle swarm algorithm which is used in the presented system. Section \ref{Sysdisc} will describe the system block diagram and then section \ref{Prepross} will explain the preprocessing step, section \ref{seg} will describe the segmentation algorithm in details. Finally, section \ref{sec:Experiments} will describe the experiments that was preformed.\\

\section{Importance of sketch recognition}
\label{sec:ImportanceOfSketchRecognition}

Scientists generally and engineers specifically express thoughts and designs using sketches. Engineers use sketches to exchange designs as a natural method of communication rather than writing or speaking. Engineers use paper and pencil design in early stages of design. The sketching process itself go thorough numerous stages. Firstly, early stages with rough drafting designs or notations which are mainly done using paper and pencil. After those primary rough sketches are edited and altered until they are stabilized and finally they are transformed into the computer.

%Design engineers need a powerful computer based system with symbol recognition to help them design, manipulate and store sketches more effectively than using only papers. Increasing the interaction between computers and users in sketch and CAD systems has been the reason for the emerging of few advanced sketch recognition systems \cite{CAD}. Sketch recognition system converts the drawings and informal graphs user draws into objects and symbols that can be used in further interaction with the system.

%People normally sketch thoughts rather than expressing them linguistically. For instance engineers mostly drafts their early designs using paper and pencil before using CAD tools.  People in general and engineers specially make use of various sketches during design process as instance organizational charts, informal drafting and formal design sketches \cite{}. The sketching process itself go thorough numerous stages. Firstly, early stages with rough drafting designs or notations which are mainly done using paper and pencil. After those primary rough sketches are edited and altered until they are stabilized and finally they are transformed into the computer.

 Paper and Pencil are preferred in design as they offer a remarkable ease in editing and altering the primary rough sketches. Current CAD systems do not give the designer the ease of altering the early designs. Designer had to sketch primary designs on paper and pencil next, the final design is transferred into computer using a CAD system menus. The process is time consuming and tiring. The need for fast design and development process is driven by the fast growing industries and technology.

 A sketch recognition and understanding system will provide a powerful tool to the designers to draw sketch as they usually draw using paper and pencil with the advantages of a computer program. Designer needs a normal sketching environment that will provide him facilities in paper interface and the modifying and simulation that is in current CAD systems.

\section{Applicability }
% talk about application like cad systems , software engineering uml diagrams ....
% meeting idea brainstorming , mind mapping string ..
% communication program  
% sketching programs ...
Application of sketch recognition are ranging from engineering device simulation and design. Drawing and sketching programs are most likely to benefit from integrating a sketch recognition system into there design environments. Other applications include object retrieval, map navigations and various engineering design systems.

@@

 Understanding hand drawn sketches may be trivial for humans but it is challenging problem when speaking about computer-based system. Most of the research made on the sketch understanding is achieved for one particular domain or for strictly a set of pre defined symbols. Further, others restrict the user the freedom in drawing which loose the main goals of paper-like-interface systems. The only few systems that have achieved the desired degree of freedom for the users are extremely computationally expensive.

Sketch recognition is divided into three steps preprocessing, segmentation and symbol recognition. The preprocessing phase captures user input stroke points and collects basic information about the stroke then proceeds to remove noise and compute basic statistical and geometrical information. In segmentation phase, the strokes are divided into a set of simple geometrical primitives or segments. In the third phase, sketch recognition, strokes and segments are clustered to formulate symbols that can be recognized by a classifier system.

The research in the area had faced few problems that will need to be handled to implement an entirely sketch recognition system. The user draws strokes in a spontaneous and ambiguous manner that intensifies the challenges. One of the problems is segmenting strokes from the way the user draw it into primitive geometric. Another major problem is the high level of ambiguities in the drawing. The next sections discripe each problem in details 
\subsection {Segmentation}
Segmentation is dividing the strokes draws into the geometric primitive's as lines, arcs, and ellipses. The strokes as mentioned before are the path of points between pen-down and pen-up events while using a stylus. The processing of those points is done after that to firstly to detect the critical points or vertices. After that the segment is classified as either line, arc, ellipse or circle. The main problem in this process is in detecting the critical and vertices which are points divide strokes into lines and arc. The user draw the lines and arc in an undefined manner which make it difficult for the program to deicide where the user had finish a line and started a new arc or vice verse; Fig [\ref {fig:strokeinterpolation}].

\begin{figure}

		\includegraphics[scale=0.8]{../../neededfiles/Figures/strokeinterpolation.eps}
	\caption[Segmentation error]{The figure shows an arc that the user draw which can be segmented in different ways as a single line or as an arc where a) The stroke users draw  b)interpretation as a single line segment  b) interpretation as an arc.}
	\label{fig:strokeinterpolation}
\end{figure}

\subsection{Ambiguity}
The user draws strokes that the computer cannot decide if it is arc or line. The nature of hand drawn sketch it self is highly ambiguous even for human eye you may see two drawn you cannot decide if it is square or rectangle.

 The ambiguities in sketch understanding are found at many levels in the system. Firstly, at the segmentation phase where the same user stroke can be interpreted as a ellipse and a circle. Symbol ambiguity is antoher problem as a symbol can have more than one interpolation. Due to the user sloppy drawing the symbols can have more than one interpolation which varies with the domain drawn in it similar problem was faced in handwriting recognition see Fig [\ref{fig:VorU}] but it was handled with knowing information about the context information which may not be the case in sketch recognition.

\begin{figure}	
	\centering
		\includegraphics[scale=0.85]{../../neededfiles/Figures/VorU.eps}
	\caption[Ambiguities Handwriting ]{The figure shows that the same letter can be interpreted differently in various contexts; first as V in drive or as U in Cut}
\label{fig:VorU}
\end{figure}

\section{Research Scope}
%rewirte this part.............................
%  research will invistgated till sketch recognition in domain of uml diagrams. 
%This research will include investigating the best method to recognize sketches.  The scope of the research will include recognizing various sketches. The methods used in this research are comparing current systems and enhancing them and adobpiting them into simple geometrical symbols.  Using well established AI methods specially \textit{Swarm Intelligence} to solve the problem. 
 This research investigats the best method to recognize sketches. The scope of the research includes recognizing various symbols drawn using single and multipule strokes. The research foucs on using various Particle Swarm Optimization (PSO) methods to segments strokes into goemetrical primitives. Sketch recognition is achieved by  Support Vector Machines (SVM) classifier. A various set of features are investigated to reach the best set of features.

\section{Thesis Overview}
This thesis is arranged such that chapter \ref{} presents a survey of different components of the
sketch recognition. In chapter \ref{}, more light is shed on the feature filter technique for dimensionality
reduction. This is in conjunction with a detailed description of the proposed methods to
enhance the feature filtering. Comparative studies among the proposed methods and stateof-
arts methods are presented in chapter 4. Finally, chapter 5 presents the conclusions of
this thesis and suggested future work.
%\subsubsection{Search space}
%\label{sec:Searchspace}
%
%
%The user drawn symbols can be having more than one interpolation. In addition, those interpolations increase exponentially with the number of strokes the user draw and as the sketch expands the number of interpolation escalate. This intensifies the search space that the program search to resolve the ambiguities and find best interpolation.
%
%\sublsubsection{Real time}
%\label{sec:Realtime}
%
%
%There are mainly three types of sketch understanding system offline \cite{Vibratory8}, online and interactive. Both online and interactive recognition like \cite{incrmentintention41, SmartSketch56} need the processing after each stroke to be real time so the user does not feel the processing.

%A picture speak a thousand words is also true for sketches. 
%Scientists generally and engineers specifically express thoughts and designs using sketches. Engineers use sketches to exchange designs as a natural method of communication rather than writing or speaking. Engineers use paper and pencil design in early stages of design. Design engineers need a powerful computer based system with symbol recognition to help them design, manipulate and store sketches more effectively than using only papers. Increasing the interaction between computers and users in sketch and CAD systems has been the reason for the emerging of few advanced sketch recognition systems \cite{CAD}. Sketch recognition system converts the drawings and informal graphs user draws into objects and symbols that can be used in further interaction with the system.

%Sketch recognition is divided into four main steps preprocessing, segmentation, symbol recognition and finally sketch understanding. The preprocessing process saves user input stroke points from the hardware and collects basic information about the stroke then proceeds to remove noise and primary data calculation. In segmentation phase, the system divides the stroke into a set of simple geometrical primitives or segments. The third phase, sketch recognition, is clustering strokes and segments to formulate symbols that can be recognized by a classifier system. Finally, the sketch understanding phase uses priori information about the sketch to support identification of symbols by excluding any possible invalid symbols. 
%the basic symbols are identified in consistent with the whole sketch. \\
%from the user information from the users converts the strokes the user draw into segments which is recognized into basic symbols that can be easily manipulated. Finally, the basic symbols are interpreted in consistent with the whole sketch. \\

>>>>

%\usepackage{graphicx}
\chapter {Introduction}
\pagenumbering{arabic}
Computer research nowadays aims at facilitating computer human interaction in every way. Pen-based interfaces give the user a pencil-paper like feeling that enhances interaction more than the current used keyboard-mouse computer interfaces. However, until now there are hardly any complete pen based computer systems. Generally, Current Pen-Based systems make use of the Pen or stylus to perform the same role as the mouse / keyboard. A minority of such systems are equipped with handwriting or digit recognition that is typically restricted for specific language.

@@

Sketches here stand for any hand drawing the user draws using digital stylus or similar device. Normally, sketches are composed of a set of strokes. The time between pen-up and pen-down events identify each stroke. The strokes are mainly the path of points, which the user path through between the pen-down and pen-up events while using the stylus or similar device.

\section{Importance of sketch recognition}
\label{sec:ImportanceOfSketchRecognition}
Scientists generally, and engineers specifically express thoughts and designs using sketches. Engineers use sketches to exchange designs as a natural method of communication rather than writing or speaking. Engineers use paper and pencil design in early stages of design. The sketching process itself goes through  numerous stages. Firstly, early stages with rough drafting designs or notations are mainly done using paper and pencil. These primary rough sketches are edited and altered until they are stabilized and finally they are transformed into the computer.

 Paper and Pencil are preferred in design as they offer a remarkable ease in editing and altering the primary rough sketches. Current CAD systems do not give the designer the ease of altering the early designs. Designer had to sketch primary designs on paper and pencil, next the final design is transferred into computer using a CAD system menus. The process is time consuming and tiring. The need for fast design and development process is driven by the fast growing industries and technology.

 A sketch recognition and understanding system will provide a powerful tool to the designers to draw sketch as they usually draw using paper and pencil with the advantages of a computer program. Designers need a normal sketching environment that will provide him facilities in paper interface and the modifying and simulation that is in current CAD systems.

\section{Applicability }
% talk about application like cad systems, software engineering uml diagrams ....
% meeting idea brainstorming , mind mapping string ..
% communication program  
% sketching programs ...
Applications of sketch recognition rang from engineering device simulation to design. Drawing and sketching programs are most likely to benefit from integrating a sketch recognition system into their design environments. Other applications include object retrieval, map navigations and various engineering design systems.

@@

Understanding hand drawn sketches may be trivial for humans but it is a challenging problem when speaking about computer-based system. Most of the research made on the sketch recognition is achieved for one particular domain or for strictly a set of pre-defined symbols\cite{Vibratory8,physicalmeaning6}. Furthermore, others restrict the user the freedom in drawing which loose the main goals of paper-like-interface systems\cite{gestureexample12,aideddesgin22,sketchinginterfaces2}. The only few systems that have achieved the desired degree of freedom for the users are extremely computationally expensive\cite{EfficientAbstract39,SketchRead2007}.

Sketch recognition is divided into three steps: preprocessing, segmentation and symbol recognition. The preprocessing phase captures user input stroke points and collects basic information about the stroke then proceeds to remove noise and compute basic statistical and geometrical information. In the segmentation phase, the strokes are divided into a set of simple geometrical primitives or segments. In the third phase, sketch recognition, strokes and segments are clustered to formulate symbols that can be recognized by a classifier system.
There are several problems in this research area that need to be handled to implement an integrated sketch recognition system. The fact that the user draws strokes in a spontaneous and ambiguous manner that intensifies the challenges. One of the main problems is segmenting spontaneous strokes drawn by the user into geometrical primitive segments. Another major problem is the high level of ambiguities in the drawing. The next sections describe each problem in details.

\subsection {Segmentation}
The goal of segmentation is dividing the strokes into the geometric primitive's as lines, arcs, and ellipses. The strokes, as mentioned before are the path of points between pen-down and pen-up events while using a stylus. Furthermore, after identifying a stroke, it should be processed to detect the critical points or vertices which divide the stroke into segments. Later, each segment is classified as line, arc, ellipse or circle. The main problem in this process is in detecting the critical points and vertices which divide strokes into lines and arcs. The user draws the lines and arcs in an undefined manner which make it difficult for the program to decide where the user had finished a line and started a new arc or vice verse;  Figure \ref {fig:strokeinterpolation}.
\begin{figure}

\begin{center}
		\includegraphics[scale=0.8]{../../neededfiles/Figures/strokeinterpolation.eps}
	\caption[Segmentation error]{Different ways for segmenting a line drawn by user. a) The stroke users draw b)interpretation as a single line segment c) interpretation as an arc.}
	%The figure shows an arc that the user draw which can be segmented in different ways as a single line or as an arc where a) The stroke users draw  b)interpretation as a single line segment b) interpretation as an arc.}
	\label{fig:strokeinterpolation}
\end{center}
\end{figure}

\subsection{Ambiguity}
Figure 1 shows an example of ambiguity, where the strokes drawn by the user can be classified in several ways. The nature of hand drawn sketch itself is highly ambiguous even for human eye. You may see two drawings you cannot decide if they are a square or a rectangle.

The ambiguities in sketch understanding are found at many levels in the system. Firstly, in the segmentation phase, a single stroke can be interpreted as an ellipse or as a circle. Symbol ambiguity is another problem as a symbol can have more than one interpolation. Due to the user's sloppy drawing the symbols can have more than one interpolation which varies with the domain drawn in it. A similar problem was faced in handwriting recognition (Figure \ref{fig:VorU}). The problem could be solved by providing context or domain information which could not be provided in most sketches. 
\begin{figure}	
	\centering
		\includegraphics[scale=0.85]{../../neededfiles/Figures/VorU.eps}
	\caption[Ambiguities Handwriting] { The same letter can be interpreted differently in various contexts; first as V in drive or as U in Cut}
\label{fig:VorU}
\end{figure}

\section{Research Scope}
%rewirte this part.............................
%  research will investigated till sketch recognition in domain of uml diagrams. 
%this research will include investigating the best method to recognize sketches.  The scope of the research will include recognizing various sketches. The methods used in this research are comparing current systems and enhancing them and adopting them into simple geometrical symbols.  Using well established AI methods specially \textit{Swarm Intelligence} to solve the problem. 
This research  presents a new method for sketch recognition. The scope of the research includes recognizing various symbols drawn using single and multiple strokes. The research compares various segmentation methods. This research used two Particle Swarm Optimization (PSO) \cite{PSOFirst} methods to segment strokes into geometrical primitives. The research also investigates various spatial, geometrical and statistical feature sets used in sketch recognition. A support Vector Machines (SVM) classifier \cite{svmintroduce} is used to classify the sketch into one of symbols from the set of previously trained symbols. This research targets the recognition of one symbol at a time. The symbol may be drawn in one or more stroke without any restriction on order, orientation or style.

%This research assumes the user draws one symbols at a time. The user can draw can draw this symbol in one stroke or in multiple stroke without any restriction on order, orientation or style.

\section{Thesis Overview}
This thesis is arranged as follows: chapter \ref{sec:Survey} presents a survey of different components of the sketch recognition. Chapter \ref{sec:ParticleSwramOptimization} gives a review on swarm intelligance and particle swarm optimization.  Chapter \ref{sec:proposedSystem} describes the proposed system and its main building blocks. The comparative study of the different techniques and their detailed results in various experiments are given in chapter \ref{sec:Evaluation}. Finally, chapter \ref{sec:DiscussionConclusion} presents the conclusions of this thesis and suggested future work.
%This thesis is arranged such that chapter \ref{} presents a survey of different components of the sketch recognition. Chapter \ref{} describe the proposed system and its main building blocks. The details description of each component is presented in Chapter \ref{}. The comparative study of different techniques with the detailed results of various experiments are detailed in chapter\ref{}. Finally, chapter \ref{} presents the conclusions of this thesis and suggested future work.
